AllMeso manuscript

### 2023-05-26 manuscript/manuscript.Rmd
Manuscript "Run all" now functional
Complete rerun overnight success

---
title: "Metabolite soup"
author: "William Kumler"
date: "10/27/2022"
output:
  word_document: default
  html_document: default
editor_options:
  chunk_output_type: inline
---

Hi Regina

```{r run analysis, eval=FALSE, echo=FALSE}
# Run untargeted analysis
knitr::purl("../untargeted/output_control.Rmd", output="../untargeted/output_script.R")
script_lines <- readLines("../untargeted/output_script.R")
polarity_set_line_idx <- grep(script_lines, pattern = "polarity <-")[1]
cruise_set_line_idx <- grep(script_lines, pattern = "cruise <-")[1]
to_write <- script_lines[-c(polarity_set_line_idx, cruise_set_line_idx)]
writeLines(to_write, "../untargeted/output_script.R")
current_dir <- getwd()
setwd("../untargeted")
for(cruise in c("FK", "MT", "MC")){
  for(polarity in c("pos", "neg")){
    print(paste("Working on", cruise, polarity))
    source("output_script.R", echo = FALSE)
  }
}

# Extract targeted results from targeted analysis

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, include = TRUE, fig.width = 6.5)

options(warn = -1)
options(dplyr.summarise.inform=F)
options(readr.num_columns = 0)
options(pillar.sigfig=7)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(RaMS))
library(scales)
library(ggdendroplot)
library(ggrepel)
library(lmPerm)

depth_levels <- c("15m", "25m", "DCMless20m", "DCMless10m", "DCM", "DCMplus10m",
                  "DCMplus20m", "175m")
filled_file_metadata <- "../metadata/made_data/filled_file_metadata.csv" %>%
  read_csv(show_col_types = FALSE) %>%
  mutate(depth=factor(depth, levels = depth_levels))

factor_levels <- c(
    "no23_um", "po4_um", "si_um", "PC_um", "PN_um", "DON_um", "DOP_um", "oxy_um",
    "chl_ug", "phaeo_ug", "alk", "dic", "ph", "theta",
    paste0(c("pro", "syn", "euk", "het"), "_abund"), "beam_atten",
    "abs_depth", "time", "lat", "sla", "local_hour"
  )
factor_labels <- c(
    "Nitrate + nitrite", "Phosphate", "Silicate", "Particulate C", "Particulate N",
    "Dissolved organic N", "Dissolved organic P", "Oxygen", "Chlorophyll a",
    "Phaeophytin", "Alkalinity", "Dissolved inorganic C", "pH", "Density",
    paste(c("Pro.", "Syn.", "Picoeuk.", "Het. bact."), "abundance"),
    "Beam attenuation",
    "Depth", "Time", "Latitude", "Sea level anomaly", "Local hour"
  )
twoline_labels <- c(
    "Nitrate + nitrite", "Phosphate", "Silicate", "Particulate C", "Particulate N",
    "Dissolved\norganic N", "Dissolved\norganic P", "Oxygen", "Chlorophyll a",
    "Phaeophytin", "Alkalinity", "Dissolved\ninorganic C", "pH", "Density",
    paste(c("Prochlorococcus", "Synechococcus", "Picoeukaryote", 
            "Heterotrophic bact."), "\nabundance"),
    "Beam attenuation",
    "Depth", "Time", "Latitude", "Sea level anomaly", "Local hour"
  )

clean_stans <- "../metadata/made_data/clean_stans.csv" %>%
  read_csv(show_col_types = FALSE) %>%
  select(compound_type, compound_name, formula, polarity)

all_feats <- read_csv("../untargeted/all_feats.csv", show_col_types = FALSE)
all_peaks <- read_csv("../untargeted/all_peaks.csv", show_col_types = FALSE)

MT_feats <- all_feats %>%
  filter(cruise=="MT") %>%
  select(-classes, -CSI_name) %>%
  rename("calc_formula"=formula) %>%
  left_join(clean_stans, by=c("compound_name", "polarity"))
MT_peaks <- MT_feats %>%
  left_join(all_peaks, by="feature") %>%
  left_join(filled_file_metadata, by="filename") %>%
  filter(samp_type=="Smp") %>%
  select(feature, norm_area, filename)
```

## Abstract

TBD

## Introduction

A major goal of marine microbial metabolomics is a global ocean model of metabolites that resolves at the molecular level, answering the question: "At any given spot and any given time, what is the concentration of a given molecule 1) within a cell and 2) in the surrounding seawater?" With such a model, we could more easily discover marine biomarkers, predict the response of marine communities to perturbations both natural and anthropogenic, and even characterize the production of cloud-forming molecules such as DMS and improve our climate models [and similar stuff, don't love these examples but like the metabolite model justification].

This is a difficult problem because the metabolome is the result of a complex network of interactions between the ocean's biology and its biogeochemistry. Gradients of light and nutrient availability create a spectrum of fundamental niches throughout the water column that are sensitive to diel, seasonal, and interannual cycles. Additionally, vertical and horizontal mixing due to upwelling/downwelling, storms, and mesoscale features alter the distribution of water masses and the dissolved molecules contained within.

These complex interactions make it difficult to predict exactly which parameters are necessary to include in the final model. Some parameters simply don't affect the outcome enough to be worth including because their effect sizes are so small that they have no biological significance. Consider an absurd parameter like "distance to nearest butterfly" that somehow obtains enormous statistical significance. If, however, it only changes values in the model on the scale of attomoles it is simply not worth calculating in the model. [Need better example here]. We can constrain the number of possible parameters by sketching out a simple ocean model, the simplest of which simply predicts an average value for every molecule at every location and time. This is, approximately, where the field of marine microbial metabolomics currently rests. For many metabolites, we have an estimate of its concentration at a single location, forcing us to extrapolate that single value vertically and horizontally throughout the ocean in the absence of other information.

With some additional oceanographic knowledge, we can predict how metabolites respond to depth. The classic Martin curve and global DOM measurements allow us to predict that the surface ocean will have the highest concentrations of metabolites and that compound abundance will generally decrease with depth. This improves our model's accuracy significantly but fails to account for the variable lability of different compounds, with some highly recalcitrant molecules having essentially conservative profiles while other molecules are assimilated extremely rapidly. This spectrum of reactivity is expected based on [blah] but has [not yet been explored.] [Mention certain compound classes? Some amino acids more labile than others?]

Even more complex models have begun to emerge in recent years informed by additional data and our understanding of ocean physics, chemistry, and biology. We have just begun to characterize the response of metabolites to gradients in time and space, covering diel variation and transects across the ocean surface. We've also begun to integrate biological information from genomic and transcriptomic studies, allowing us to better resolve the theoretical metabolome as biology responds to shifts in light intensity and nutrient availability. The model shown below in black represents one potential graph of these interactions and the role of metabolites as mediators of the interaction between organisms.

```{r fig.height=3}
DiagrammeR::grViz(diagram = "digraph dot {
  graph [rankdir = LR]

  tab1 [label = '@@1', pos='0, 1!']
  tab2 [label = '@@2', pos='0, 0!']
  tab3 [label = '@@3', pos='2, 0.5!']
  tab4 [label = '@@4', pos='2,-0.5!']
  tab5 [label = '@@5', pos='4, 1!']
  tab6 [label = '@@6', pos='4,-1!']
  tab7 [label = '@@7', pos='4, 0!']
  tab8 [label = '@@8', pos='0,-1!', color='red', fontcolor='red', style='dashed']

  tab1 -> tab3;
  tab2 -> tab3;
  tab2 -> tab4;
  tab3 -> tab5;
  tab4 -> tab5;
  tab4 -> tab6;
  tab5 -> tab7;
  tab6 -> tab7;
  tab7 -> tab5;
  tab7 -> tab6;
  tab8 -> tab4 [color='red', style='dashed']
}

  [1]: 'Time of day'
  [2]: 'Depth'    
  [3]: 'Light'
  [4]: 'Nutrients'
  [5]: 'Phototrophs'
  [6]: 'Heterotrophs'
  [7]: 'Metabolites'
  [8]: 'Sea level anomaly'
  ", engine="neato")
```

However, the model as is predicts a uniform concentration of nutrients at a given depth. This is not observed in the real world due to mixing and the vertical movement of water masses in response to mesoscale eddies. These features can be identified by satellites as anomalies in sea surface height and reflect the fluctuations of density surfaces within the water column.

Cyclonic eddies, those circulating counterclockwise in the northern hemisphere and clockwise in the southern, bring the main pycnocline closer to the surface of the ocean and introduce extra nutrients to the sunlit layer of the ocean upon formation. Anticyclonic eddies have the opposite effect, depressing isopycnal surfaces and essentially piling up surface water in regions of positive sea level anomaly (SLA). These eddies can increase the concentrations of nutrients at the base of the photic zone by more than an order of magnitude and are known to affect phytoplankton species composition [and other species?].

The magnitude of this effect on the metabolome is currently unknown, potentially representing a major source of variation in the marine metabolite model. If metabolites are as sensitive to SLA as nutrients are, then the eddy state (shown in red in the figure) will be a critical parameter to include in the marine metabolomic model described above.

Certain metabolites have been studied previously with respect to SLA, most notably chlorophyll [and pheophytin] [for its role in estimating the state of the phytoplankton community]. Fluorometric measurements of chlorophyll are regularly taken during CTD casts and the deep chlorophyll maximum (DCM) represents an important water column feature for oceanographic sampling [and biological blah]. The DCM has been noted previously to fluctuate up to [XX] meters in response to fluctuations in sea surface height, indicating that at least one metabolite is sensitive to eddy state in a biologically significant way. However, the amplitude of this fluctuation is much less dramatic than that of the nutricline or isopycnal surfaces because it represents a compromise between the constant influx of light from above and the varying nutrient field, with light essentially serving as a stabilizing force for the DCM.

In this paper, we describe the metabolome's variability with respect to depth and sea level anomaly with the goal of constraining these parameters in a [future? hypothetical?] global ocean model of marine metabolites. The driving questions we answer are listed below:

1. Is the metabolome as a whole significantly affected by depth, time of day, and sea level anomaly? Which of these parameters causes the most variability in the metabolome and is therefore most important to include in a global ocean model?

2. How do individual metabolite depth profiles vary in the upper ocean? Can we distinguish labile compounds that are readily degraded from those that are more recalcitrant? Are there chemical or biological similarities between metabolites that respond similarly to changes in depth?

3. Which metabolites are sensitive to eddy state? Do they respond more like nutrients and isopycnal surfaces, or are they stabilized by light availability as the DCM is? Are there any metabolites that can be found more abundantly or even exclusively in cyclonic (or anticyclonic) eddies, perhaps serving as biomarkers for the biological community within?

These questions have not been answered previously because measuring compound abundances in the ocean is difficult. Marine metabolites typically have concentrations in the nanomolar-picomolar range and the small, polar molecules in particular are difficult to separate from the enormously more abundant salt ions. Fortunately, recent advances in mass spectrometry and liquid chromatography have produced methods that allow the measurement of these low-abundance molecules with sufficient accuracy and rapidity for this type of analysis. Here, we use a high-resolution accurate-mass instrument coupled to a hydrophilic interaction column to rapidly assay all the small molecules in a given sample.

Additionally, advances in data analysis and computational power have made untargeted mass-spectrometry increasingly viable as a technique for detecting and quantifying molecules that exist beyond the limited suite traditionally pursued in environmental metabolomics. Unlike targeted MS, which only analyzes a select few compounds known in advance, untargeted metabolomics is data-driven and can pick out signals produced by compounds that are unknown or unannotated. This is a significant advantage in environments such as the ocean where comprehensive databases have yet to be established and provides a way to make mass-spectrometry analysis perfectly reproducible via scripting.

## Methods

### Cruise information

To answer the questions established above, samples were collected from the North Pacific Subtropical Gyre near Station ALOHA during two research cruises that targeted strong mesoscale eddy features. The first cruise, KM1709, occurred during June and July 2017 and traversed an area between 28 °N, 157° W and 23 °N, 158.5 °W. For this cruise, an eddy dipole off the coast of Hawaii was detected using sea-level anomaly (SLA) satellite data and targeted for both a transect across the cyclonic and anticyclonic poles of the eddy dipole as well as two long-term Lagrangian stations at the center of each eddy. The cyclonic pole of the eddy had a maximum negative SLA anomaly of -20 centimeters and the anticyclonic center reached 24 centimeters. The transect samples were taken at various times of day as the ship transected the dipole, and the eddy centers were all sampled between 5 and 8 pm. (See Supplemental Table XX for cruise data) The second cruise occurred during March and April of 2018 and only sampled from the center of each eddy. This Falkor cruise (cruise number FK180310) also traveled between two eddies of different polarities, a cyclonic eddy at 22 °N 156 °W and an anticyclonic eddy at 24.5 °N and 161 °W. The cyclonic eddy was sampled first and had a maximum negative SLA difference of -12 cm while the anticyclonic eddy was sampled one day later and had a maximum SLA of 20 cm. Samples on this cruise were taken twice a day, once at 6AM and once at 6PM. (Supp table XX)

### Environmental data collection

For collaborators to fill in. Need details on every bit of environmental metadata:

  - Bottle 
    - Chlorophyll/phaeophytin
    - Nutrients
    - PC/PN (note outliers)
    - DON/DOP
    - Alkalinity/DIC/pH
    - Theta? This is corrected pressure, right? Not just pressure?
    - Oxygen
    - Pro/Syn/Picoeuk/Het
  - Beam attenuation
  - PAR?
  - SLA (pulled from CMAP)

Environmental parameters were interpolated to the locations at which metabolite samples were taken using a [bilinear algorithm].

### Metabolite sample collection

Samples were obtained using the onboard CTD rosette to collect water from 15 meters, the deep chlorophyll maximum, and 175 meters during KM1709 eddy transect, from the DCM ±10 and ±20 meters during the KM1709 eddy center sampling, and from 25 meters and the DCM during FK180310. The DCM was determined visually from fluorometer data (XX fluorometer from XX company) during the CTD downcast and Niskin bottles were tripped during the return trip to the surface. 10 liters of seawater were collected in each Niskin and each depth was sampled in triplicate by firing multiple bottles simultaneously. Samples were brought to the surface and decanted into prewashed (3x with DI, 3x with sampled seawater) polycarbonate bottles for filtration. Samples were filtered by peristaltic pump onto 142mm 0.2 µm Durapore [Omnipore?] filters [part number] held by polycarbonate filter holders [part number] on a Masterflex [part number] tubing line. Pressures were kept as low as possible while still producing a reasonable rate of flow through the filter, approximately 250 mL per minute. Samples were then removed from the filter holder using solvent-washed [ethanol?] tweezers and placed into pre-combusted aluminum foil packets that were then flash-frozen in liquid nitrogen before being stored at -80 °C until extraction. A methodological blank was also collected by running filtrate through a new filter and then treated identically to the samples.

### Sample extraction

Extraction followed a modified Bligh & Dyer approach as detailed in Boysen et al. (2018). Briefly, filters were added to PTFE centrifuge tubes with a 1:1 mix of 100 µm and 400 µm silica beads, approximately 2mL -20 °C Optima-grade DCM, and approximately 3mL -20 °C 1:1 methanol/water solution (both also Optima-grade). Extraction standards were added after this step, a table of which is available as SI Table XX. The samples were then bead-beaten three times, followed by triplicate washes with fresh methanol/water mixture. Samples were then dried down under clean N2 gas and kept warm using a Fisher-Scientific Reacti-Therm module. Dried aqueous fractions were re-dissolved in 380 µL of Optima-grade water and amended with 20 µL isotope-labeled injection standards, also documented in SI Table XX. The dry organic fraction was reconstituted into 380 µL 90:10 DCM:toluene [right?] rather than Boysen et al.’s 1:1 water:acetonitrile. Both fractions were also finally syringe-filtered [part number?] to remove any potential clogging material. The aqueous fraction was then aliquot into an HPLC vials one for injection on the HILIC column. The organic fraction was not run. [Were the samples themselves diluted? I think yes?] A pooled sample was created by combining 20 µL of each sample into the same HPLC vial, and a 1:1 dilution with water half-strength sample was aliquot from that to assess matrix effects and obscuring variation per Boysen et al. (2018). Also run alongside the environmental samples were two mixes of authentic standards in water and in an aliquot of the pooled sample at a variety of concentrations, available in SI Table XX, for quality control, annotation, and absolute concentration calculations. HPLC vials containing the samples were frozen at -80 °C until thawing shortly before injection.

### HPLC-MS methods

[Verbatim from CH_Ingalls_Lab_LC_Methods from Katherine’s MW submission] A SeQuant ZIC-pHILIC column (5 um particle size, 2.1 mm x 150 mm, from Millipore) was used with 10 mM ammonium carbonate in 85:15 acetonitrile to water (Solvent A) and 10 mM ammonium carbonate in 85:15 water to acetonitrile (Solvent B) at a flow rate of 0.15 mL/min. The column was held at 100% A for 2 minutes, ramped to 64% B over 18 minutes, ramped to 100% B over 1 minute, held at 100% B for 5 minutes, and equilibrated at 100% A for 25 minutes (50 minutes total). The column was maintained at 30 °C. The injection volume was 2 µL for samples and standard mixes. When starting a batch, the column was equilibrated at the starting conditions for at least 30 minutes. To improve the performance of the HILIC column, we maintained the same injection volume, kept the instrument running water blanks between samples as necessary, and injected standards in a representative matrix (the pooled sample) in addition to standards in water. After each batch, the column was flushed with 10 mM ammonium carbonate in 85:15 water to acetonitrile for 20 to 30 minutes.

Metabolomic data was collected on a Thermo Q Exactive HF hybrid Orbitrap (QE) mass spectrometer.

The capillary and auxiliary gas heater temperatures were maintained at 320°C and 100°C, respectively. The S-lens RF level was kept at 65 [units], but the H-ESI voltage was set to 3.3 kV instead and sheath gas, auxiliary gas, and sweep gas flow rates were decreased to 16, 3, and 1 [units], respectively. Polarity switching was again used but with a scan range of 60 to 900 m/z and a resolution of 60,000. Calibration was performed every 3-4 days using the [Thermo Calibration Mix (part number?)] at a target mass of 200 m/z. DDA data was collected by fragmenting the pooled samples with [details??? Same in HILIC and RP?]. All files were then converted to an open-source mzML format via Proteowizard’s msConvert tool [version number??] with flags and filters specified in Supplemental Table XX [centroiding, pos/neg separation].

### Targeted metabolomic methods

Targeted compounds were manually integrated using Skyline [version number] and MSDIAL [version number] with parameters specified in Supplemental Table XX. Skyline was used for the KM1709 and FK180310 eddy center samples, while MSDIAL was used for the KM1709 eddy transect data. Raw peak areas were then normalized to their best-matched internal standard per Boysen et al. (2018). The coefficient of variance (CV) threshold used to choose a suitable match was X.X, as determined by looking at the internal standards normalized to each other and selecting the minimum CV that did not introduce new variation, per the tools in the B-MIS normalization GitHub repository (https://github.com/IngallsLabUW/B-MIS-normalization). Normalized peaks were then calibrated via a single-point standard curve as determined uniquely for each compound from the authentic samples used to convert normalized peak area into moles per µL injected and scaled to the amount of seawater filtered to provide a final estimate for each compound of moles per liter of seawater filtered.

### Untargeted metabolomic methods

Untargeted compounds were initially analyzed using the xcms package [version] from the Bioconductor project [version] in the R programming language [version]. Parameters for import and peakpicking can be found in Supplemental Table XX, and the functions readMSdata and findPeaks were used to produce a list of raw peak areas. However, because the default xcms algorithm has several flaws in its signal-to-noise ratio (SNR) estimation, [cite Myers & Du 2017] custom code was written to remove many of these features likely to be noise. To estimate SNR more robustly, the raw m/z, retention time, and intensity data was extracted from the mzML files using the R package RaMS [version] and subset for each peak using the mzmin, mzmax, rtmin, and rtmax values produced by xcms. Each peak was matched against a Gaussian curve with varying degrees of skew [provide values!] to account for peak tailing effects, and the curve with the smallest residuals was chosen for further analysis. The raw data was then subtracted from this ideal curve to estimate the residual variation (noise) present in the peak, and the maximum height of the peak was divided by the standard deviation of these residuals to estimate SNR. The advantage of this method is that it can be used on peaks that do not have many data points outside of the peak itself, a feature common to HILIC data. This is unlike the existing algorithms in xcms and MzMine2 which require a large window of background points to correctly calculate SNR. This new SNR was used to calculate an overall “peak quality” index calculated as

$$ SNR * peak\ correlation^4*max\ intensity$$

where peak correlation is the Pearson’s correlation coefficient calculated between the Gaussian curve and the real data and max intensity is the height of the peak. Peaks with a quality score less than 20 were excluded from the remainder of the analysis. [Defend the choice of 20]. This list of peaks was then fed back into the default XCMSnExp object and followed by the default functions for alignment and correspondence, parameters for which can also be found in Supplemental Table XX. Finally, a custom version of xcms’ default fillPeaks algorithm was used due to bugs in the source code that prevented the appropriate parameters from being used.

After the curated list of peaks was produced, isotopes and adducts were removed. Isotopes and adducts were detected by again accessing the raw data via the R package RaMS [version number] and subsetting between rtmin and rtmax, but instead the new m/z limits were calculated by assuming the detected peak was the isotope or adduct of another and that assumed peak’s mass was used for the subsetting. The raw data from that m/z offset was then compared via Pearson correlation to the raw data of the original peak for a raw similarity comparison. Additionally, the Pearson correlation of the peak areas for each feature across multiple files was calculated as an orthogonal metric of similarity. If the raw similarity correlation was above 0.8 and the correlation between files was above 0.99, the feature was assumed to be an adduct or isotope and removed from downstream analysis as well. The list of adducts and isotopes checked for each feature can be found in Supplemental Table XX.

While untargeted data analysis makes it challenging to provide absolute quantification for given molecules, the authentic standards run for the targeted analysis made it possible to annotate a portion of the untargeted peaks as known compounds at a confidence level of 1 in the MSI reporting criteria (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3772505/). This was accomplished by comparing each known metabolite to the data collected from the authentic standards in water data. The highest annotation certainty was assigned to those compounds with isotopically labeled internal standards – if a known metabolite had an internal standard, the m/z value for both the compound and its isotopologue were calculated. All peaks for each m/z value were collected and those with the closest retention time were assumed to be the isotope and its internal standard, removing them from further annotation consideration. The next highest certainty was calculated using the multiple authentic standard mixes we ran alongside the samples themselves. To discriminate between isomers of known compounds, these mixes were designed to alternate between isomers in retention time such that the distance between sequential peaks in a given mix was maximized. This allows us to calculate a between-mixes ratio that can then be compared to the known mix in which a given compound was detected. If the feature appears in the correct mix (defined as having an area in the correct mix one full standard deviation larger than that of the areas detected in the incorrect mix), it is assigned the metabolite name and removed from further annotation steps. The next highest certainty was calculated for known metabolites that had multiple features at the same m/z value, typically isomers. If the number of features matched the number of authentic standards expected at a given m/z, they were annotated by rank-order retention time matching (i.e. the first feature was assumed to be the authentic standard with the lowest retention time, followed by the second feature assumed to be the authentic standard with the second-lowest retention time, etc.). Finally, peaks were annotated using retention time similarity to the expected value in the standards list and ultimately annotated as the peak with the largest area if all the above failed. [Revisit the above to explain interactions and the whole if/else thing]. Compounds that were multiply assigned (either one feature assigned to multiple authentic standards or one standard assigned to multiple features) were manually deconvoluted.

Untargeted metabolites were also normalized to a best-matched internal standard (IS) as defined above in the targeted metabolomics section. However, the IS were manually integrated using Skyline because even small outliers in integration area significantly affected the downstream analysis after compounds were normalized to them. The cutoffs used for untargeted data analysis were [higher/lower] than those in the targeted suite, with an “already good” CV of 0.X and a minimum improvement of 0.X.

Chemical formulae were estimated from each metabolite's m/z ratio and isotope envelope information. We implemented this estimation using three separate approaches, using first a set of custom functions, then the R package Rdisop version X.XX.XXX, and finally SIRIUS version X.XX.XXX. To create isotope envelopes for detected features, we wrote code very similar to that used to remove the adducts and isotopes above, but rather than assuming the feature was an isotope or adduct we simply looked at the m/z values assuming that the feature was the [M+H]$^+$ adduct in positive mode and the [M-H]$^-$ adduct in negative mode. The adducts and isotopes tested can also be found in Supplemental Table XX. After calculating the m/z values for each feature, the raw data was again accessed using RaMS and the trapezoidal Riemann sum was calculated between the minimum retention time and the maximum retention time for the feature across each file and adduct or isotope. Raw data with a Pearson correlation above 0.75 and a correlation between files of 0.9 or greater was assumed to be an isotope of the base peak.

For the custom calculations, these values were used to estimate the ratio of heavy to light abundance. A Type I linear model was fit to each feature’s isotope ratios and the slope of the line was then used to estimate the number of atoms contained in each formula using the calculations provided in Supplemental Methods XX. This custom method provided a way to constrain the results produced by Rdisop and SIRIUS. For those functions, the file-specific isotope envelope had to be collapsed into a feature-based format. The average monoisotopic area was used for the base peak and multiplied by the slope of the line from the linear model to obtain an average isotope ratio across the multiple files. Next, the Rdisop’s “decomposeIsotopes” function was used with a ppm error of 5 if the m/z ratio was above 200 or a fixed tolerance of 0.001 Da otherwise. The potential formula candidates output from this function were then filtered by presence/absence in Pubchem while relaxing the formula to contain one more or fewer hydrogen atoms than initially estimated to account for [M+H]+ and [M-H]- adducts. Finally, SIRIUS command-line functions were used to provide a third orthogonal estimate of molecular formula. The “decomp” function was passed the m/z value for each peak and constrained with a ppm error of 5 and an element list of CHNOPS. Only formulae that appeared in both Rdisop and SIRIUS and also had an elemental breakdown matching the custom calculations were kept for further analysis.

The above untargeted process was applied separately to each combination of column type, polarity, and sample run. This resulted in six unique peak lists that required correspondence across analyses. The six HILIC runs were correspondedby selecting the feature with the largest area and features from the other runs near the largest feature in *m/z* and retention time were selected, with the features from each other run closest in area and retention time according to log$_2$ retention time and log$_{10}$ average area identified as most likely candidates and grouped. Those features were then removed from consideration and the correspondence was repeated for the next largest peak until no features remained ungrouped. Features that had no corresponding feature in other runs became their own group. Finally, grouped features that contained a targeted compound were used to identify precision and accuracy of the grouping. A manual retention time correction of 60 seconds was applied to the features detected in the FK180310 cruise after peak visualization showed a consistent offset of approximately that distance in the data.

### Statistics

Fourteen environmental variables were measured during the eddy transect alongside the metabolite samples ([Figure|Table XX]). We confirmed that these parameters were able to uniquely characterize differences due to depth in the eddy transect with both univariate and multivariate methods. First, we tested for differences due to sampling depth using Welch's ANOVA (@welch1951), chosen because variances could not be assumed equal for most measurements and corrected for multiple comparisons using Holm's method (@holm1979). Next, we distributed the samples into a two-dimensional space using non-metric multidimensional scaling (NMDS) as implemented in R's vegan package with the metaMDS function. Two dimensions were chosen because one dimension would not provide sufficient explanatory power and the stress value was below 0.1 when two dimensions were used, rendering a third was unnecessary. For this NMDS analysis, each environmental variable was z-score scaled using base R's scale function to have a mean of zero and unit standard deviation, thereby making each parameter unit-invariant and allowing us to disable autotransformation. Finally, the environmental vectors were fit onto this ordination using vegan's envfit function, allowing us to determine the direction in the ordination space in which each vector changes most quickly and estimate the significance of each fitted factor using permutation.

To assess the significance of depth, eddy polarity, and diel variability on the metabolome as a whole, we performed a Permutational Multivariate Analysis of Variance (PERMANOVA) using R's vegan package (version XX). For this analysis, the QC'd and BMIS'd peak areas were ranked using R's rank function with default ties.method argument of "average" to place all metabolites on the same scale. Discrete SLA bins were created by categorizing each station as cyclonic if the corrected SLA was less than -5cm, anticyclonic if greater than 5cm, or neither if between these endpoints (@barone). Discrete time categories were created by binning sample times into morning, midday, evening, and night if the sample was collected between 3:30am and 9:30am, 9:30am and 3:30pm, 3:30pm and 9:30pm, or 9:30pm and 3:30am, respectively. These bins were chosen [arbitrarily a priori, see if changes happen when using the MAHALO bins instead per Josh's recommendation]. Depth was already discretized by the depth at which the sample was taken: 15 meters, the DCM, and 175 meters. The adonis2 function was used with a model formula consisting of a ranked metabolite data matrix as described above as the LHS and a RHS of either no interactions, pairwise interactions, or all interactions. Significance was assessed marginally by setting the "by" argument to "margin" to avoid the potentially confounding effect of assessing terms sequentially, and pairwise distances were calculated using the Manhattan distance metric. Nonmetric multidimensional scaling (NMDS) was performed with the metaMDS function, also in the vegan package, again using the Manhattan distance metric. Three dimensions were chosen because two-dimensional solutions consistently exceeded a stress of 0.15. Autotransform was disabled because the data have already been normalized via ranking. The Manhattan distance metric was chosen for its more intuitive performance in higher dimensions, but the results did not change [very much? significantly?] when the Euclidean distance metric was used instead [double check this].

To characterize the response of metabolites to depth, we ran a Kruskal-Wallis on each compound/parameter followed by Dunn's post-hoc test for pairwise comparisons as implemented by the rstatix's package dunn_test function with no p-value adjustment if the Kruskal-Wallis was significant at an alpha level of 0.05 after adjusting p-values to control the false discovery rate using R's p.adjust function. The Kruskal-Wallis test was chosen instead of the traditional parametric ANOVA because compounds and parameters could not be assumed to have equal variance or normal error distributions, and Dunn's test was chosen as the post-hoc for the same reason. The Dunn test p-values were not corrected for multiple comparisons because [blah]. Compounds were then classified into one of twelve categories by the sign, significance, and magnitude of the post-hoc test results, as well as a null category (no detectable response with depth) and a fourteenth category for compounds with a significant Kruskal-Wallis but for which Dunn's test was unable to significantly discriminate between depth pairs. These categories [emerge from permuting boxplot thingys and there's twelve for REASONS and it's not arbitrary] [I need to describe how I assigned the categories in more detail but haven't figured out how to do that yet - maybe a table?].

To test the hypothesis that compounds most abundant at the surface would have the highest C:N ratios and compounds most abundant deeper in the ocean would have the lowest C:N ratios, we [blah blah blah, used the vertical axis of the clockplot as a proxy for "most abundant at surface" vs "most abundant at depth" and ran a regression of C:N ratio vs y-axis coordinate on 1) only known metabs, 2) all metabs with estimated molecular formulae].

To measure the effect of SLA at each sampled depth, we ran a simple linear regression (Model I) on each metabolite at each depth individually, using SLA as the predictor variable and rank-normalized metabolite abundance as the dependent. Separating the data set in this way allowed us to estimate the impact of SLA on metabolite abundance [without the confounding effect of variable responses due to depth]. Compounds were then categorized into a "positive trend with SLA" category if they had a significantly positive slope, a "negative trend with SLA" if they were significantly negative, or a null category of no significant trend if SLA did not have a detectable effect on metabolite abundance.

We assessed the relative importance of sea level anomaly (SLA) on the metabolome by using the separate set of samples collected from the center of the eddies on the KM1709 cruise. These depths were chosen to represent a large, but not enormous, amount of variability due to niche partitioning in the water column. We then compared metabolite abundances between the two eddies using a simple Student's t-test and grouped all five unique depths together.

We measured the similarity of the metabolome by calculating a distance matrix between each metabolite that appeared in all three data sets after normalizing each to the maximum value observed for that metabolite within each data set. We then calculated Pearson's correlation coefficient between these three distance matrices, allowing us to see how well each metabolome corresponded to each other. To compare the robustness of individual compounds, we ran permutational tests for differences due to SLA on all three data sets; a t-test for the Falkor and eddy center data and a regression for the eddy transect data. We then identified compounds that responded significantly to SLA at an alpha level of 0.05 across different data sets.

All statistics were performed in the R programming environment, [version].

## Results

### Environmental data

Measured values for all fourteen environmental variables are shown below in [Table|Figure] XX. Much of the variance observed in these values is explained by the collection of samples at three discrete depths: 15 meters below the surface, the deep chlorophyll maximum, and 175 meters. DCM depth was not a fixed distance from the surface but instead fluctuated with sea level anomaly in response to nutrient availability such that the DCM samples varied in their absolute depth from 100 to 133 meters.

```{r metadata boxplots & table}
distinct_MT_metadata <- filled_file_metadata %>%
  filter(cruise=="MT") %>%
  mutate(basename=str_remove(untripl, "180821_Smp_")) %>%
  select(basename, depth, sla:last_col()) %>%
  group_by(basename) %>%
  summarise(depth=unique(depth), across(-depth, mean)) %>%
  select(-PAR, -PAR_perc, -ph, -dic, -alk, -oxy_um, -theta) %>%
  filter(!is.na(basename))

# If we decide to use a figure (boxplots of values by depth)
distinct_MT_metadata %>%
  mutate(depth=factor(depth, levels = c("175m", "DCM", "15m"))) %>%
  select(basename, depth, sla, ends_with("abund"),
         ends_with("um"), ends_with("ug"), beam_atten) %>%
  pivot_longer(-c(basename, depth, sla)) %>%
  mutate(name=factor(name, levels = factor_levels, labels = twoline_labels)) %>%
  filter(!is.na(value)) %>%
  ggplot() +
  geom_boxplot(aes(x=value, y=depth)) +
  facet_wrap(~name, scales="free_x", nrow=3) +
  theme(axis.title = element_blank())

>>>>

---
title: "AllMeso manuscript"
author: "William Kumler"
date: "`r Sys.Date()`"
output: html_document
---

```{r Run analysis}
if(!file.exists("../metadata/filled_file_metadata.csv")){
  setwd("../metadata")
  knitr::purl("metadata_control.Rmd")
  source("metadata_control.R")
  file.remove("metadata_control.R")
  setwd("../manuscript")
}
if(!file.exists("../untargeted/all_peaks.csv")){
  setwd("../untargeted")
  data_iters <- c("FK_pos", "FK_neg", "MS_pos", "MS_neg")
  knitr::purl("untargeted_control.Rmd")
  for(cruise in c("FK", "MS")){
    for(polarity in c("pos", "neg")){
      source("untargeted_control.R")
    }
  }
  file.remove("untargeted_control.R")
  map(data_iters, function(output_dir){
    comb_peaks <- read_csv(paste0(output_dir, "_output/clean_peak_data.csv")) %>%
      mutate(temp=output_dir) %>%
      separate(temp, into = c("cruise", "polarity"), sep = "_") %>%
      mutate(feature=str_replace(feature, "FT", paste0(toupper(output_dir), "_")))
  }) %>%
    bind_rows() %>%
    write_csv("all_peaks.csv")
  map(data_iters, function(output_dir){
    comb_peaks <- read_csv(paste0(output_dir, "_output/clean_feat_data.csv")) %>%
      mutate(temp=output_dir) %>%
      separate(temp, into = c("cruise", "polarity"), sep = "_") %>%
      mutate(feature=str_replace(feature, "FT", paste0(toupper(output_dir), "_")))
  }) %>%
    bind_rows() %>%
    write_csv("all_feats.csv")
  setwd("../manuscript")
}
if(!file.exists("../targeted/all_peaks.csv")){
  setwd("../targeted")
  knitr::purl("../targeted/targeted_control.Rmd")
  source("targeted_control.R")
  file.remove("targeted_control.R")
  setwd("../manuscript")
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

filled_file_metadata <- read_csv("../metadata/filled_file_metadata.csv")
clean_stans <- read_csv("../metadata/clean_stans.csv")
untarg_feats <- read_csv("../untargeted/all_feats.csv")
untarg_peaks <- read_csv("../untargeted/all_peaks.csv")
targ_peaks <- read_csv("../targeted/all_peaks.csv")
```
